---
title: "Want to sink a boat?"
format: html
---



Setup
================================================================================

- Hit the broom in your global environment pane to clear out your memory.
- Then run this setup chunk to import the Titanic data into memory.

```{r}
#| label: setup
#| echo: false
#| warning: false

library(tidymodels)
library(tidyverse)
options(scipen = 999)

titanic <-
  read_csv("data/titanic-train-clean.csv") |>
  mutate(
    pclass = factor(pclass),
    child = relevel(
      as.factor(if_else(child == 0, "No", "Yes")),
      ref = "Yes"
    ),
    died = !survived,
    died = relevel(
      as.factor(if_else(died == TRUE, "died", "survived")),
      ref = "survived"
    )
  ) |>
  select(-survived)
```



Task 01
================================================================================

- **Question:** What is the log odds ratio of third-class passengers to first
  class passengers?
- **Answer:** 1.67

- **Question:** What is the log odds ratio of second-class passengers to first
  class passengers?
- **Answer:** 0.64

```{r}

titanic |>
  mutate(
    .died = died == "died",
    .survived = died == "survived"
  ) |>
  group_by(pclass) |>
  summarize(
    died = sum(.died),
    survived = sum(.survived)
  ) |>
  mutate(
    odds = died / survived
  ) |>
  summarise(
    or_2 = odds[pclass == 2] / odds[pclass == 1],
    log_odds_2 = log(or_2),
    or_3 = odds[pclass == 3] / odds[pclass == 1],
    log_odds_3 = log(or_3)
  )

```



Task 02
================================================================================

- **Question:** What is the coefficient of pclass3?
- **Answer:** 1.67

- **Question:** In a few sentences, tell me what this coefficient tells us about
  the odds of a third-class passenger surviving.
- **Answer:** exp(1.6704) For every one first-class passenger who died, over 
  five and one/third, third-class passengers died.
    - If they write this as 5 1/3, 5.3, or even five, they get credit for their
      answer.

- **Question:** What does the intercept tell us? This is a one sentence answer,
  but you will have to see if you can deduce where this number comes from.
- **Answer:** The intercept is the log odds of first-class passengers.

```{r}

glm_pclass <-
  logistic_reg() |>
  fit(
    died ~ pclass,
    data = titanic
  )
glm_pclass

```



Task 03
================================================================================

- **Question**: What is the accuracy of your model from Task 02?
- **Answer:** .68 OR 67.90.
- Helpful Hints:
    - You only need to enter this out to TWO significant digits.

```{r}

titanic_glm_pclass <-
  augment(
    glm_pclass,
    new_data = titanic
  )

accuracy(
  data = titanic_glm_pclass,
  truth = died,
  estimate = .pred_class
)

```



Task 04
================================================================================

- **Question:** What is the accuracy of your new model, `glm_lab`?
- **Answer:** .79 or 79.12

- **Question:** Is this model more accurate, less accurate, or about the same
  compared to model `glm_pclass`?
- **Answer:** It is more accurate.

- **Question:** Of the metrics I have taught you, sensitivity is the lowest. In
  a single sentence, explain what the sensitivity metric is telling you about 
  this model.
- **Answer:** Of all passengers who died, this model only correctly identifies
  69% of them.
    - They must make reference to the value of sensitivity as 69% or .69 to get 
      full credit here.

```{r}
glm_lab <-
  logistic_reg() |>
  fit(
    died ~ pclass + child + sex,
    data = titanic
  )
glm_lab

titanic_glm_lab <-
  augment(
    glm_lab,
    titanic
  )


conf_mat(titanic_glm_lab, truth = died, estimate = .pred_class)
accuracy(titanic_glm_lab, truth = died, estimate = .pred_class)
sensitivity(titanic_glm_lab, truth = died, estimate = .pred_class)
specificity(titanic_glm_lab, truth = died, estimate = .pred_class)

```



Task 05
================================================================================

- **Question:** Based on everything you have seen in this lecture and lab, do
  you think the adage "women and children first" is an accurate description for 
  who was saved when Titanic sank?
- **Answer:** Yes and no. Women and children were more likely to survive, but
  passenger class was also highly predictive. It's like they left that that part
  out.
    - This is an experimental question and I'm not sure what we are going to get.
    - We can discuss how to grade this one together.