---
title: "Want to sink a boat?"
format: html
---



Setup
================================================================================

- Hit the broom in your global environment pane to clear out your memory.
- Then run this setup chunk to import the Titanic data into memory.

```{r}
#| label: setup
#| echo: false
#| warning: false

library(tidymodels)
library(tidyverse)
options(scipen = 999)

titanic <-
  read_csv("data/titanic-train-clean.csv") |>
  mutate(
    pclass = factor(pclass),
    child = relevel(
      as.factor(if_else(child == 0, "No", "Yes")),
      ref = "Yes"
    ),
    died = !survived,
    died = relevel(
      as.factor(if_else(died == TRUE, "died", "survived")),
      ref = "survived"
    )
  ) |>
  select(-survived)
```



Task 01
================================================================================

Another famous facet of the Titanic tragedy is that first and second class
passengers were more likely to survive than third class passengers. Create a 
model which predicts death as a function of passenger class (`pclass`).

I will go ahead and tell you that first class passengers were more likely to
survive than any other passenger class. Therefore, we can calculate the odds 
ratio for second and third class passengers, compared to first class passengers.

Using the odds of first class passengers as dying, calculate the odds ratio for
second and third class passengers.

- **Question:** What is the log odds ratio of third class passengers to first
  class passengers?
- **Answer:** 

- **Question:** What is the log odds ratio of second class passengers to first
  class passengers?
- **Answer:** 

- Helpful Hints:
    - Use the column `pclass` to identify passenger class.
    - Use the odds for first class passengers in the denominator.
    - Use the odds for second class passengers in the numerator.
    - Calculate the log of this odds ratio.
    - The log of the odds ratio is very sensitive to rounding. But to make it 
      easier to enter into Canvas, I will only ask for the first two significant
      digits. From there, follow standard rounding rules.
    - However, when calculating the `log()` or `exp()` of a number, use all
      available digits.

```{r}

## YOUR CODE HERE!!!

```



Task 02
================================================================================

Create a logistic regression model which predicts death as a function of 
passenger class (`pclass`). This model should return an intercept, and 
coefficients for pclass2 and pclass3.

- **Question:** What is the coefficient of pclass3?
- **Answer:** 

- **Question:** In a few sentences, tell me what this coefficient tells us about
  the odds of a third class passenger surviving.
- **Answer:** 

- **Question:** What does the intercept tell us? This is a once sentence answer,
  but you will have to see if you can deduce where this number comes from.
- **Answer:** 
- Helpful Hints:
    - Your work on Task 01 is relevant for figuring this out!
    - There's one group missing . . . 

```{r}

## YOUR CODE HERE!!!

```



Task 03
================================================================================

- **Question**: What is the accuracy of your model from Task 02?
- **Answer:** 
- Helpful Hints:
    - You only need to enter this out to TWO significant digits.

```{r}

## YOUR CODE HERE!!!

```



Task 04
================================================================================

In the lecture, we built a model using `sex` and we built a model using `child`.
These were "simple" logistic regression models because the probability of dying
was calculated as a function of a single feature. As with linear regression, we
can build multiple logistic regression models which use multiple features.

Build a model called glm_lab which predicts the probability of death as a 
function of:

- Passenger Class `pclass`
- Child/Adult `child`
- Sex `sex`

The order in which you list the features does not matter to the end result.
After building the model, review the confusion matrix, accuracy, sensitivity,
and specificity of this new model.

- **Question:** What is the accuracy of your new model, `glm_lab`?
- **Answer:** 

- **Question:** Is this model more accurate, less accurate, or about the same
  compared to model `glm_pclass`?
- **Answer:** 

- **Question:** Of the metrics I have taught you, sensitivity is the lowest. In
  a single sentence, explain what the sensitivity metric is telling you about 
  this model.
- **Answer:** 
- Helpful Hints: 
    - Use the sensitivity percent calculated by R as part of your answer.
    - If you don't, you won't get full credit.


```{r}

## YOUR CODE HERE!!!

```



Task 05
================================================================================

Please signal to me when you get to this task.

- **Question:** Based on everything you have seen in this lecture and lab, do
  you think the adage "women and children first" is an accurate description for 
  who was saved when Titanic sank?
- **Answer:**
- Helpful Hints:
    - Use the numbers we calculated in the lecture and in the lab to support
      your answer.
    - Please don't write a five-paragraph theme. A few sentences, supported by 
      data is sufficient.